Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-14 aistats.tex
minor modifs

{\bf Estimators.}
In general, we want to estimate $\nat$ from some data $\mD = (x_1, \dots, x_n)$.
For this purpose, not only is $f$ convex, but it even enjoys a closed form maximum-likelihood estimate (MLE)
\begin{align}
	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
	\label{eq:defMLE}
\end{align}
This rule is known as moment matching.
Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).

{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality:
how close is it from some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish two ways of doing that.
{\bf 1)} distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \&p(x)>0$. 
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction} , the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
Hence the fundamental problem:
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

We have asymptotic solutions (\S\ref{ssec:asymptote}),
or when $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
or close to quadratic (see \S\ref{ssec:quadratic}),
but a  general solution for finite $n$ remains elusive.
In this paper, we review partial solutions, and give some ideas on how to solve it.

{\bf Stochastic optimization} offers an interesting perspective on this question.
Consider the problem
\alignn{
	\min_{\theta\in \Theta} f(\theta)
	\label{eq:optimization_problem}
}
solved by $\nat^*\in \Theta$.
Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
\alignn{
	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} }
	\label{eq:suboptimalityKL}
}
Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we try using existing optimization convergence rates for SMD to get the upper bound we seek.
Unfortunately, none of the current analysis apply, which frames another open problem for the analysis of SMD.

{\bf Expected Ouctomes.}
A solution to~\eqref{problem} would quantify the importance of the prior in MAP, in particular in the few sample regime, for instance for Gaussians $\cN(m,\sigma^2)$. 
More generally it could enable stochastic optimization to tackle a wide class of barrier objectives\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}.
A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
It could also help in assessing the impact of alternative forms of regularization (prior) for these models.

{\bf Contributions.}
After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance})
	\item By identifying MAP and stochastic mirror descent (SMD), we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
\end{itemize}

{\bf Notation.}
$X$ and $T=T(X)$ are random variables, while $x$ is a sample.
$n$ is the number of samples and $d= \dim(T)$.
$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.

@@

{\bf Duality.}
The log-partition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
Given that $\logpart$ is strictly convex, its Hessian is positive definite and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both are linked and represent the same distribution.

We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
\begin{align}
	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
\end{align}
which is the common notion of \textit{entropy} in information theory.
By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$, giving
\aligns{
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
}
%\begin{figure}[t]
%	\centering
%	\includegraphics{duality}
%	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
%	}
%	\label{fig:duality}
%\end{figure}%
%
%Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
{\bf The Bregman Divergence} induced by $\logpart$,
measures the discrepancy between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
    \label{eq:defBregman}
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$.
In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
%Finally, it is equal to the divergence of its convex conjugate with switched arguments,
%\begin{align}
%	\bregman (\nat ; \nat_0)
%    = \bregmanconj ( \meanp_0 ; \meanp) \; .
%    \label{eq:bregman_switch}
%\end{align}

@@

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
\textcolor{red}{The explanation here is not clear enough, or too concise} 
Now we can match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\aligns{
	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
	\conj(\m) = - \half \log(\m)  \; ,
}
up to constants.
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m_*; \m_n)
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
	\bregman( \nat_n; \nat_* )
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in \cref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptote~\eqref{eq:asymptote} .
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
Note that this inequality actually holds for the symmetrized Bregman $\cB(a,b) + \cB(b,a)$, which explains why it loses a factor 2 compared to the asymptote~\eqref{eq:asymptote}, but we leave this to the appendix for clarity.
Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth
This correlates well with our numerical observations (cf App.\ref{app:gaussian-variance}).

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.
	}
	\label{fig:thales}
\end{figure*}

\subsection{Gaussian}
\label{ssec:gaussian}
The ubiquitous Gaussian $\cN(m,\sigma^2)$ actually offers a highly non-trivial example for Problem 1.
Its log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
With sufficient statistic $T(x)=(x, x^2)$, 
the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$. 
$\cM$ and $\Theta$ are delimited in grey in \cref{fig:thales}.
Given these parameters, log-partition and entropy are
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2)
}
up to constants.

 Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.

@@

\paragraph{Illustrations.}
We plot the bias-variance decompositions for $\cN(0,\sigma^2)$ (Figure~\ref{fig:variance_decomposition}) and  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:thales} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) . \rlp{Tune figures.}\textcolor{red}{Not super clear, what is the "characters featured in these decompositions?"}

@@

\textbf{Caveat :} all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging in the primal.

@@

\section{Related Work}

\paragraph{Online Learning}
Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
\[
R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
\]
where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$. Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule. Unfortunately however, their rate are worse than the offline MAP or MLE algorithms. For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$. From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$. Also, \citet{dasgupta2007online} prove the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$. This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.

>>>>

{\bf Estimators.}
In general, we want to estimate $\nat$ from some data $\mD = (x_1, \dots, x_n)$.
For this purpose, not only is $f$ convex, but it even enjoys a closed form maximum-likelihood estimate (MLE)
\begin{align}
	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
	\label{eq:defMLE}
\end{align}
This rule is known as moment matching.
Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
The parameter $\theta$ can also be estimated with online-learning techniques, but they are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.

{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality, i.e., its closeness to some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish here two ways: \textit{disntance in parameter space} and \textit{distance between distributions}.
{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \&p(x)>0$. 
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
Hence the fundamental problem:
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

We have already asymptotic solutions (\S\ref{ssec:asymptote}) when $\logpart$ is quadratic (e.g. $X$ is Gaussian with known variance)
or close to quadratic (\S\ref{ssec:quadratic}).
However, a  general solution for finite $n$ remains elusive.
In this paper, we review partial solutions, and give ideas on how to solve it.

{\bf Stochastic optimization} offers an interesting perspective on this question.
Consider the problem
\alignn{
	\min_{\theta\in \Theta} f(\theta)\,,
	\label{eq:optimization_problem}
}
solved by $\nat^*\in \Theta$.
Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
\alignn{
	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} }
	\label{eq:suboptimalityKL}
}
Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we try using existing optimization convergence rates for SMD to get the upper bound we seek.
Unfortunately, none of the current analyses apply, which frames another open problem for the analysis of SMD.

{\bf Expected Ouctomes.}
A solution to~\eqref{problem} would quantify the importance of the prior in MAP, in particular in the few sample regime.%, for instance for Gaussians $\cN(m,\sigma^2)$. 
Also, it could enable stochastic optimization to tackle a wide class of barrier objectives\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}.
A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
It could also help in assessing the impact of alternative forms of regularization (prior) for these models.

{\bf Contributions.}
After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
	\item By linking MAP and SMD, we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
\end{itemize}

{\bf Notation.}
$X$ and $T=T(X)$ are random variables, $x$ is a sample, $n$ is the number of samples and $d= \dim(T)$.
$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.

@@

{\bf Duality.}
The log-partition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \\
    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0,
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
Given that $\logpart$ is strictly convex, its Hessian is positive definite and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both are linked and represent the same distribution.

We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
\begin{align}
	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
\end{align}
which is the common notion of \textit{entropy} in information theory.
By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$, giving
\aligns{
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
}
%\begin{figure}[t]
%	\centering
%	\includegraphics{duality}
%	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
%	}
%	\label{fig:duality}
%\end{figure}%
%
%Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
{\bf The Bregman Divergence} induced by $\logpart$ measures the discrepancy between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
    \label{eq:defBregman}
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$.
In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
%Finally, it is equal to the divergence of its convex conjugate with switched arguments,
%\begin{align}
%	\bregman (\nat ; \nat_0)
%    = \bregmanconj ( \meanp_0 ; \meanp) \; .
%    \label{eq:bregman_switch}
%\end{align}

@@

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its density reads
\[
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
\]
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
\textcolor{red}{The explanation here is not clear enough, or too concise} 
Now we can match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\begin{align}
	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
	\conj(\m) &= - \half \log(\m) + \half \log\frac{\pi}{2} - \half  \; .
\end{align}
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m_*; \m_n)
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
	\bregman( \nat_n; \nat_* )
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in \cref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptote~\eqref{eq:asymptote} .
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
Note that this inequality actually holds for the symmetrized Bregman $\cB(a,b) + \cB(b,a)$, which explains why it loses a factor 2 compared to the asymptote~\eqref{eq:asymptote}, but we leave this to the appendix for clarity.
Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth, which correlates well with numerical observations (cf App.\ref{app:gaussian-variance}).
%For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
%Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
	\label{fig:bias-variance-numerical}
\end{figure*}

\subsection{Gaussian}
\label{ssec:gaussian}
The simple Gaussian $\cN(m,\sigma^2)$ is highly non-trivial. Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.

@@

\paragraph{Illustrations.}
We plot the bias-variance decompositions for $\cN(0,\sigma^2)$ (Figure~\ref{fig:variance_decomposition}) and  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:bias-variance-numerical} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) . \rlp{Tune figures.}\textcolor{red}{Not super clear, what is the "characters featured in these decompositions?"}

@@

Caveat : all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.

@@

%\section{Related Work}

%\paragraph{Online Learning}
%Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
%\[
%R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
%\]
%where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$. Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule. Unfortunately however, their rate are worse than the offline MAP or MLE algorithms. For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$. From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$. Also, \citet{dasgupta2007online} prove the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$. This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
